Using Model Checking to Analyze the System Behavior of the LHC Production Grid

### 2013-02-04 ElsevierFGCS_singleColumn_old.tex
Keep old version before review.

The development of DIRAC started in 2002 as a system for production of simulation
data that would serve to verify physics theory, aspects of the LHCb detector design, as
well as to optimize physics algorithms. It gradually evolved into a complete community
grid solution for data and job management, based on a general-purpose framework
that can be reused by other communities besides LHCb. Today, it covers all major
LHCb tasks starting with the raw data transfer from the experiment's detector to
the grid storage, several steps of data processing, up to the final user
analysis. 
It provides a concurrent use of over 60K CPUs and 10M file replicas distributed over
tens of grid sites, provided by the WLCG computing grid \cite{Bird:2011zz}. The community of users has grown to several hundreds, loading the 
LHC grid resources with up to 30K simultaneously running jobs (Fig.~\ref{fig:LHCb-resource-usage}, top) and over 100000 jobs executed per day, during peak processing periods.
About 10PB of data is stored and visible in the LHCb file catalogs, with daily data transfers among storage units reaching 3Gbps (Fig.~\ref{fig:LHCb-resource-usage}, bottom).
\begin{figure*}[tp!]
  \centering
  {\label{fig:Job-throughput}\includegraphics[width=0.7\linewidth]{./graphics/Figure1_1.png}}    
  \vfill
  \vspace{14pt}
{\label{fig:Data-transfers}\includegraphics[width =0.7\linewidth]{./graphics/Figure1_2.png}}
  \caption{LHCb resource usage}
  \label{fig:LHCb-resource-usage}
\end{figure*}
Python was chosen as the implementation language, since it enables
rapid prototyping and development of new features. DIRAC follows the Service
Oriented Architecture (SOA) paradigm, accompanied by a network of lightweight
distributed agents which animate the system. Its main components are depicted in
Fig.~\ref{fig:DIRAC-Arch}.

\begin{figure}[tp]
\includegraphics[width=0.5\textwidth] {./graphics/Figure1.png}
\centering
\caption{DIRAC Architecture overview}
\label{fig:DIRAC-Arch}
\end{figure}

The \textit{services} are passive components that react to requests from their clients,
possibly soliciting other services in order to fulfill the requests. They run as
permanent processes deployed on a number of high-availability hosts (VO-boxes)
at CERN, and store the dynamic system state information in database
repositories. The user interfaces, agents or running jobs can act as clients
placing the requests to DIRAC's services.

@@

The driving force of DIRAC is the Workload Management System (WMS). Taking into
account the heterogeneous, dynamic, and high-latency nature of the distributed computing environment, a
\textit{Pilot Job paradigm}, illustrated in Fig.~\ref{fig:DIRAC-WMS}, 
\begin{figure}[bp]
\includegraphics[width=0.7\textwidth]{./graphics/Figure2.png}
\centering
\caption{DIRAC Workload Management System \cite{DIRAC_pilot_WMS}}
\label{fig:DIRAC-WMS}
\end{figure}
was chosen as an efficient way to
implement a pull scheduling mechanism.  Pilot jobs are simply resource
reservation processes without an actual payload defined a priori. They are
submitted to the grid worker nodes with the aim of checking the sanity of the
operational environment just before pulling and executing the real payload. This
hides the fragility of the underlying resources and increases the job success
rate, from the perspective of end users. Centralized \textit{Task Queues} contain all the
pending jobs, organized in groups according to their requirements. This enables
efficient application of job priorities, enforcing fair-share policies and
quotas, which are decided within the LHCb community, without the need of any
effort from the grid sites.  Originally, the DIRAC WMS submitted pilot jobs using 
the native gLite middleware tools, but this method was shown to be suboptimal. The gLite resource brokers
are centralized components, and as such they do not have the capacity and flexibility to cope with the load of 
a community such as LHCb. Using multiple resource brokers, rather than a single one, was not the solution:
all brokers use the same site state information and would always submit jobs to the most suitable one globally, causing overloads to
certain sites while underloading others very often. In addition, the gLite information system was not reactive enough to propagate
changes in site status information to the brokers. With the wide deployment of the CREAM CE service \cite{Aiftimiei2010654}, DIRAC's WMS switched
to submitting pilot jobs directly to sites. By interrogating the Computing Element status directly, the system reactiveness
improved dramatically, making sites compete with each other for jobs and have a faster job turnaround.

@@

\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure7_new.png}

@@

on the CacheReplicas memory process, which resulted in only 6 interesting states for this process alone.

The clustering
allowed us to see precisely which chain of actions by the concurrent agents
advanced the state of the CacheReplicas entity. In fact, in the modeling phase we discovered the first problematic
SMS behavior, blocking the progress of staging
tasks (and as a result the progress of jobs). The origin of the deadlock had been a
trivial human logic flaw introduced in coding. At a particular circumstance,
when a ``New'' task enters the system \textbf{and} all associated replicas
are already ``Staged'' (possibly by other jobs requiring the same input files),
its status would immediately be promoted to ``Done''.
Such tasks would never be picked up by SMS and called-back to the appropriate job, 
since the agent responsible for this final step would only look for ``Staged'' tasks.
A proper state machine implementation
instead of hardcoding states in SQL queries can prevent such bugs.
\begin{figure}[bp!]
\vspace{-10 pt}
\includegraphics[width=0.57\linewidth,keepaspectratio=true]{./graphics/Figure8.png}
\centering
\caption{State-space visualisation of the SMS with ltsview}
\label{fig:LTSviewVisualisation}
\end{figure}%123478

>>>>

The development of DIRAC started in 2002 as a system for production of simulation
data that would serve to verify physics theory, aspects of the LHCb detector design, as
well as to optimize physics algorithms. It gradually evolved into a complete community
grid solution for data and job management, based on a general-purpose framework
that can be reused by other communities besides LHCb. Today, it covers all major
LHCb tasks starting with the raw data transfer from the experiment's detector to
the grid storage, several steps of data processing, up to the final user
analysis. 
It provides a concurrent use of over 60K CPUs and 10M file replicas distributed over
tens of grid sites. The community of users has grown to several hundreds, loading the 
LHC grid resources with up to 50K simultaneously running jobs during peak processing periods.
Python was chosen as the implementation language, since it enables
rapid prototyping and development of new features. DIRAC follows the Service
Oriented Architecture (SOA) paradigm, accompanied by a network of lightweight
distributed agents which animate the system. Its main components are depicted in
Fig.~\ref{fig:DIRAC-Arch}.

\begin{figure}[tp]
\includegraphics[width=0.5\textwidth] {./graphics/Figure1.png}
\centering
\caption{DIRAC Architecture overview}
\label{fig:DIRAC-Arch}
\end{figure}
The \textit{services} are passive components that react to requests from their clients,
possibly soliciting other services in order to fulfill the requests. They run as
permanent processes deployed on a number of high-availability hosts (VO-boxes)
at CERN, and store the dynamic system state information in database
repositories. The user interfaces, agents or running jobs can act as clients
placing the requests to DIRAC's services.

@@

The driving force of DIRAC is the Workload Management System (WMS). Taking into
account the heterogeneous, dynamic, and high-latency nature of the distributed computing environment, a
\textit{Pilot Job paradigm}, illustrated in Fig.~\ref{fig:DIRAC-WMS}, 
\begin{figure}[bp]
\includegraphics[width=0.7\textwidth]{./graphics/Figure2.png}
\centering
\caption{DIRAC Workload Management System \cite{DIRAC_pilot_WMS}}
\label{fig:DIRAC-WMS}
\end{figure}
was chosen as an efficient way to
implement a pull scheduling mechanism.  Pilot jobs are simply resource
reservation processes without an actual payload defined a priori. They are
submitted to the grid worker nodes with the aim of checking the sanity of the
operational environment just before pulling and executing the real payload. This
hides the fragility of the underlying resources and increases the job success
rate, from the perspective of end users. Centralized \textit{Task Queues} contain all the
pending jobs, organized in groups according to their requirements. This enables
efficient application of job priorities, enforcing fair-share policies and
quotas, which are decided within the LHCb community, without the need of any
effort from the grid sites.

@@

\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure7.png}

@@

on the CacheReplicas memory process, which resulted in only 7 interesting states for this process alone.

The clustering
allowed us to see precisely which chain of actions by the concurrent agents
advanced the state of the CacheReplicas entity. In fact, in the modeling phase we discovered the first problematic
SMS behavior, blocking the progress of staging
tasks (and as a result the progress of jobs). The origin of the deadlock had been a
trivial human logic flaw introduced in coding. At a particular circumstance,
when a ``New'' task enters the system \textbf{and} all associated replicas
are already ``Staged'' (possibly by other jobs requiring the same input files),
its status would immediately be promoted to ``Done''.
Such tasks would never be picked up by SMS and called-back to the appropriate job, 
since the agent responsible for this final step would only look for ``Staged'' tasks.
A proper state machine implementation
instead of hardcoding states in SQL queries can prevent such bugs.
\begin{figure}[bp!]
\vspace{-10 pt}
\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure8.png}
\centering
\caption{State-space visualisation of the SMS with ltsview}
\label{fig:LTSviewVisualisation}
\end{figure}%123478
